The 2018 heatwave and its implications on ozone induced damage on vegetation in a subarctic climate

### 2020-05-05 VegOzone.tex
Added the corresponding citeations for the reports from Norway and Sweden.

Following a report by the Norwegian Meteorological institute ({\bf TODO: Citation https://www.met.no/publikasjoner/met-info/met-info-2019}), the summer of 2018 was the warmest and driest ever recorded in eastern, western, and southern Norway. In the north (including Finnmark), it was amongst the warmest on record - favorable conditions for boundary layer ozone formation.

@@

A total of 2079 forest fires of different size were registered in Norway in 2018, twice as many as in the preceding years 2016/17 ({\bf TODO: Cite https://www.dsb.no/reportasjearkiv/brannstatistikk-2018/}, last accessed April 2020). In Sweden, about 500 fires had been reported (five times more than in a usual summer) and an estimated total of $25000\,\unit{hectare}$ burned down in the course of the most extensive fires in central Sweden (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) in younger history ({\bf TODO: How to cite governmental report of Sweden? https://www.regeringen.se/rattsliga-dokument/statens-offentliga-utredningar/2019/02/sou-20197/ and report therein}).

@@

Within the scope of our project to assess probable effects of midnight sun conditions on ozone uptake by plants and resultant enhanced susceptibility to ozone inflicted damage, we conducted ambient air ozone concentration measurements at Svanhovd in the growing seasons of 2018/19 in collaboration with NILU. 
In 2018, we have observed ozone damage on semi-natural vegetation (clover) and crops (tobacco) in the ozone garden at NIBIO Environment Centre Svanhovd. In contrast, no such damage was found on the clovers in 2019, although it was found on the sensitive tobacco cultivar. This indicates that the vegetation may have been more affected by ozone in 2018 than in 2019.
The respective observed average ozone concentration in 2018/19, however, did not differ significantly by statistical means.

@@

{\bf TODO: Some more (technical) details about ozone and atmospheric monitoring at Svanhovd. $\rightarrow$ Sverre?}

\emph{Comment: We refrain from doing a full cross-evaluation for these meteorological fields, since this has been done elsewhere. It's somewhat clear that this is a job of the operators in the monitoring network.} {\bf TODO: Find a citation for this -> Met NOR report!.}.

@@

Ozone concentrations \chem{[O_3]} measured in $2\,\unit{m}$ height above ground are averaged hourly. As can be seen in Fig.~\ref{fig:data_svanvik_2018_2019}a), \chem{[O_3]} peaks in spring (April/May) and reaches its minimum in late summer (July/August). The spring peak has not been captured completely in 2019, for data acquisition started later than in 2018. In summer 2018 (June--August), high ozone concentrations ($\chem{[O_3]} > 40\,\unit{ppb}$) were recorded 50 times. The highest summer ozone concentration ($\chem{[O_3]} = 50.2\,\unit{ppb}$) was measured on July 25. This coincides with the period of the most extensive forest fires in central Swedish (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) which occurred from July 12--29 and destroyed about $18000\,\unit{hectare}$ of forest{\bf TODO: Cite chapter 11 Swedish government report.}. {\bf TODO: Do we need a corresponding \chem{CO_2} emission here?} However, due to the above mentioned data acquisition problems, we missed most of the corresponding ozone data for this event and most likely also the peak ozone concentration. In contrast, ozone concentration only rose 18 times above the threshold of $40\,\unit{ppb}$ during the summer of 2019.

@@

\subsubsection{Surface ozone observations}
\label{subsubsec:ebas}
All surface ozone observations used in this study are available from the EBAS database operated by NILU (\href{http://ebas.nilu.no/}{ebas.nilu.no}). In Table~\ref{tab:ebas_obs}, we give an overview over the exact locations and time range of operational data taking. All stations with available long term observations are located on higher ground than Svanhovd which will result in higher surface ozone concentrations in comparison as ozone abundance, in general, decreases with decreasing altitude in the absence of local precursors. {\bf TODO: Citation!} {\bf TODO: The following has to be confirmed and elaborated on! Ane?} The vegetation at all sites is similar and consists mainly of pine forests, birch, and heath.

@@

\begin{figure*}[t]
  \includegraphics[width=12cm]{ozone_timeseries_fenoscandic_obs.png}
  \caption{Time series of ozone observations in northern Fennoscandia (Tab.~\ref{tab:ebas_obs}). Data taken from EBAS. The hatched areas indicate time periods with insufficient quality control. \bf{TODO: citation NILU ozone report 200x}.}
  \label{fig:ozone_timesseries_fenoscandic_obs}
\end{figure*}

@@

Details of reanalysis data sets referred to in this work are listed in Table~\ref{tab:ozone_rea}.
There are two global reanalysis products available at the European Centre for Medium-Range Weather Forecasts (ECMWF) that include atmospheric tracers, in particular ozone (MACC and CAMSRA). The temporal as well as spatial resolution of these reanalysis products is rather coarse: 3-hourly and $0.75^\circ\times 0.75^\circ$ or roughly $29.3\,\unit{km}\times 83.4\,\unit{km}$ at the location of Svanvik, respectively. In addition, higher resolution surface ozone reanalysis for whole Europe is available from Copernicus service for regional air quality control ({\bf TODO: From May 31st on through the CAMS Atmosphere Data Store (ADS)}). The regional air quality model reanalysis ensemble (CAMS regional Ensemble) is based on nine European state-of-the-art numerical air quality models ({\bf TODO: Reference -  https://atmosphere.copernicus.eu/documentation-regional-systems, last accessed April 2020}). The ensemble mean is at fairly high spatial and temporal resolution compared to the global reanalysis from ECMWF: $0.1^\circ\times 0.1^\circ$ or roughly $3.9\,\unit{km}\times 11.1\,\unit{km}$ and 1-hourly, respectively. The covered time periods differ and no data are available for before the turn of the millennium. The CAMSRA and CAMS Regional Air Quality ensemble mean are available in near real time, but only former covers a period of sufficient length for climate analysis (2003 -- present). All reanalyses use different but, at that time, latest versions of the operational weather forecast system (OpenIFS) from ECMWF. They differ substantially in the assimilated observational ozone data. The MACC reanalysis assimilates only satellite derived tropospheric column ozone (TC), while CAMSRA also includes ozone profiles from satellite retrievals. Most importantly for our study, in situ observations from ozone surface station networks are only assimilated in the CAMS Regional Ensemble. The quality with respect to observations in northern Fennoscandia will be discussed in Section~\ref{subsubsec:clim_ozone}. Without too much foreshadowing, the results prevented us from fetching the whole CAMSRA data.

@@

The density distribution for Svanvik and the respective \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) display a similar pattern as the northern Fennoscandic climatology. The spring peak apparently occurs slightly earlier (in the beginning of April). As Svanvik is located at lower altitude than the other sites, the earlier ozone peak may point to an interplay with ozone uptake by vegetation. {\bf TODO: The data from SeNorge seems to point to a later onset of greening, than what I interfered from OpenIFS data. This means that the beginning of the growing season is not a good measure for the termination of the ozone spring peak. This does not exclude the early start of pine forest photosynthesis as a driver. Coniferous forest starts to photosynthesis earlier than what the growing season begin would mark. I remember Håkan or Per Erik(?) talking about the \chem{CO_2} uptake and phenology of coniferous trees - but still have not found a related publication? Is there a way to confirm this? Actual data for the three sites? Please correct if wrong! Ane? Aud?} The average ozone concentration at Svanvik is lower than at the other sites, as already mentioned, which is in league with a decrease in tropospheric ozone with decreasing altitude. In July--September, ozone is occasionally almost completely depleted during night time.

@@

There are many probable explanations for the discovered differences in the climatologies derived from the three reanalyses. The effective changes from the MACC reanalysis to CAMSRA have been reported and discussed by {\bf TODO: Find paper and citation.} on a global level. Amongst others, an assimilation of ozone profiles from satellite retrieval and an upgraded ozone chemistry scheme have lead to the enhanced performance of latter. The higher spatial resolution of the regional air quality models used in the CAMS Regional Ensemble compared to the global reanalyses does capture more localized depletion and peak episodes in \chem{[O_3]} better and enhances the capability of the models with respect to daily and seasonal cycles in ozone {\bf TODO: Any citable source for this assumption?}. Latter is, of course, also affected by upgrades in the OpenIFS used as meteorological forcing. However, we can assume that the main driver for the strikingly different results are the assimilated observational ozone data. Satellite retrievals resolve \chem{[O_3]} at the surface rather poorly and hence do not constrain the global models well enough {\bf TODO: Any citable source for this?}.

>>>>

Following a report by the Norwegian Meteorological institute \citep{MetNOR2019}, the summer of 2018 was the warmest and driest ever recorded in eastern, western, and southern Norway. In the north (including Finnmark), it was amongst the warmest on record - favorable conditions for boundary layer ozone formation.

@@

A total of 2079 forest fires of different size were registered in Norway in 2018, twice as many as in the preceding years 2016/17 \citep{DSB2019}, last accessed April 2020). In Sweden, about 500 fires had been reported (five times more than in a usual summer) and an estimated total of $25000\,\unit{hectare}$ burned down in the course of the most extensive fires in central Sweden (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) in younger history \citep{SOU2019}.

@@

Within the scope of our project to assess probable effects of midnight sun conditions on ozone uptake by plants and resultant enhanced susceptibility to ozone inflicted damage, we conducted ambient air ozone concentration measurements at Svanhovd in the growing seasons of 2018/19 in collaboration with NILU. 
In 2018, we have observed ozone damage on semi-natural vegetation (clover) and crops (tobacco) in the ozone garden at NIBIO Environment Centre Svanhovd. In contrast, no such damage was found on the clovers in 2019, although it was found on the sensitive tobacco cultivar. This indicates that the vegetation may have been more affected by ozone in 2018 than in 2019.
The respective observed average ozone concentration in 2018/19, however, did not differ significantly by statistical means. {\bf TODO: Citation NILU ozone report 2019 - high AOT40 in 2018 \citep{NILU2019}.}

@@

{\bf TODO: Some more (technical) details about ozone and atmospheric monitoring at Svanhovd? $\rightarrow$ Sverre?}

\emph{Comment: We refrain from doing a full cross-evaluation for these meteorological fields, since this has been done elsewhere. It's somewhat clear that this is a job of the operators in the monitoring network \citep{MetNOR2019}.}

@@

Ozone concentrations \chem{[O_3]} measured in $2\,\unit{m}$ height above ground are averaged hourly. As can be seen in Fig.~\ref{fig:data_svanvik_2018_2019}a), \chem{[O_3]} peaks in spring (April/May) and reaches its minimum in late summer (July/August). The spring peak has not been captured completely in 2019, for data acquisition started later than in 2018. In summer 2018 (June--August), high ozone concentrations ($\chem{[O_3]} > 40\,\unit{ppb}$) were recorded 50 times. The highest summer ozone concentration ($\chem{[O_3]} = 50.2\,\unit{ppb}$) was measured on July 25. This coincides with the period of the most extensive forest fires in central Swedish (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) which occurred from July 12--29 and destroyed about $18000\,\unit{hectare}$ of forest \citep{SOU2019}. {\bf TODO: Do we need a corresponding \chem{CO_2} emission here?} However, due to the above mentioned data acquisition problems, we missed most of the corresponding ozone data for this event and most likely also the peak ozone concentration. In contrast, ozone concentration only rose 18 times above the threshold of $40\,\unit{ppb}$ during the summer of 2019.

@@

\subsubsection{Surface ozone observations}
\label{subsubsec:ebas}
All surface ozone observations used in this study are available from the EBAS database operated by NILU (\href{http://ebas.nilu.no/}{ebas.nilu.no}). In Table~\ref{tab:ebas_obs}, we give an overview over the exact locations and time range of operational data taking. All stations with available long term observations are located on higher ground than Svanhovd which will result in higher surface ozone concentrations in comparison as ozone abundance, in general, decreases with decreasing altitude in the absence of local precursors. {\bf TODO: Citation see references in \citet{AB:Klingberg2009}!} {\bf TODO: The following has to be confirmed and elaborated on! Ane?} The vegetation at all sites is similar and consists mainly of pine forests, birch, and heath.

@@

\begin{figure*}[t]
  \includegraphics[width=12cm]{ozone_timeseries_fenoscandic_obs.png}
  \caption{Time series of ozone observations in northern Fennoscandia (Tab.~\ref{tab:ebas_obs}). Data taken from EBAS. The hatched areas indicate time periods with insufficient quality control \citep{NILU2003}.}
  \label{fig:ozone_timesseries_fenoscandic_obs}
\end{figure*}

@@

Details of reanalysis data sets referred to in this work are listed in Table~\ref{tab:ozone_rea}.
There are two global reanalysis products available at the European Centre for Medium-Range Weather Forecasts (ECMWF) that include atmospheric tracers, in particular ozone (MACC and CAMSRA) \citep{ACP:Inness2013, ACP:Inness2019}. The temporal as well as spatial resolution of these reanalysis products is rather coarse: 3-hourly and $0.75^\circ\times 0.75^\circ$ or roughly $29.3\,\unit{km}\times 83.4\,\unit{km}$ at the location of Svanvik, respectively. In addition, higher resolution surface ozone reanalysis for whole Europe is available from Copernicus service for regional air quality control ({\bf TODO: From May 31st on through the CAMS Atmosphere Data Store (ADS)}). The regional air quality model reanalysis ensemble (CAMS regional Ensemble) is based on nine European state-of-the-art numerical air quality models ({\bf TODO: Reference -  https://atmosphere.copernicus.eu/documentation-regional-systems, last accessed April 2020}). The ensemble mean is at fairly high spatial and temporal resolution compared to the global reanalysis from ECMWF: $0.1^\circ\times 0.1^\circ$ or roughly $3.9\,\unit{km}\times 11.1\,\unit{km}$ and 1-hourly, respectively. The covered time periods differ and no data are available for before the turn of the millennium. The CAMSRA and CAMS Regional Air Quality ensemble mean are available in near real time, but only former covers a period of sufficient length for climate analysis (2003 -- present). All reanalyses use different but, at that time, latest versions of the operational weather forecast system (OpenIFS) from ECMWF. They differ substantially in the assimilated observational ozone data. The MACC reanalysis assimilates only satellite derived tropospheric column ozone (TC), while CAMSRA also includes ozone profiles from satellite retrievals. Most importantly for our study, in situ observations from ozone surface station networks are only assimilated in the CAMS Regional Ensemble. The quality with respect to observations in northern Fennoscandia will be discussed in Section~\ref{subsubsec:clim_ozone}. Without too much foreshadowing, the results prevented us from fetching the whole CAMSRA data.

@@

The density distribution for Svanvik and the respective \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) display a similar pattern as the northern Fennoscandic climatology. The spring peak apparently occurs slightly earlier (in the beginning of April). As Svanvik is located at lower altitude than the other sites, the earlier ozone peak may point to an interplay with ozone uptake by vegetation. {\bf TODO: The data from SeNorge seems to point to a later onset of greening, than what I interfered from OpenIFS data. This means that the beginning of the growing season is not a good measure for the termination of the ozone spring peak. This does not exclude the early start of pine forest photosynthesis as a driver. Coniferous forest starts to photosynthesis earlier than what the growing season begin would mark. \chem{CO_2} uptake and phenology of coniferous trees \citep{TB:Kolari2007} starts as early as doy 100, acclimates to leaf temperature history - photosynthesis drops at around doy 200--250.} The average ozone concentration at Svanvik is lower than at the other sites, as already mentioned, which is in league with a decrease in tropospheric ozone with decreasing altitude. In July--September, ozone is occasionally almost completely depleted during night time.

@@

There are many probable explanations for the discovered differences in the climatologies derived from the three reanalyses. The effective changes from the MACC reanalysis to CAMSRA have been reported and discussed by \citet{ACP:Inness2019} on a global level. Amongst others, an assimilation of ozone profiles from satellite retrieval and an upgraded ozone chemistry scheme have lead to the enhanced performance of latter. The higher spatial resolution of the regional air quality models used in the CAMS Regional Ensemble compared to the global reanalyses does capture more localized depletion and peak episodes in \chem{[O_3]} better and enhances the capability of the models with respect to daily and seasonal cycles in ozone {\bf TODO: Any citable source for this assumption?}. Latter is, of course, also affected by upgrades in the OpenIFS used as meteorological forcing. However, we can assume that the main driver for the strikingly different results are the assimilated observational ozone data. Satellite retrievals resolve \chem{[O_3]} at the surface rather poorly and hence do not constrain the global models well enough {\bf TODO: Any citable source for this?}.
